SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-11-06 2017-Xudong-SMT/ActaInformatica.tex
Acta Inf. paper: Add explanation on page 22.
minor changes

\begin{example}[The ``Enable'' Operator]\label{example:EnableOperator}
\ERIC{In Fig. \ref{schema:enable-pnets} we shall give two possible pNet  encodings of the ``Enable'' operator respectively state-oriented and data-oriented. 
The first one is called $EnableState$, and uses a controller $CState$ with three simple control actions $l, r, \delta$ in it to synchronise the behaviour of $P$ and $Q$. Here we use a simple action algebra, containing two constructors $\delta(x)$ and $acc(x)$, for any possible data type of the variable $x$, corresponding to the LOTOS statements \texttt{exit} and \texttt{accept}. To construct the algebra presentation, we extend the triple $\APres=\langle\mathit{Sorts},\mathit{Constrs},\mathit{Ops}\rangle$ shown in Table \ref{table:BIPalgebra} with four new action constant constructors $l, r, \delta, acc$.}

@@

\ERIC{The $EnableData$ pNet will use the same algebra presentation.}\QIN{The main difference between $EnableState$ and $EnableData$ lies in the construction of their controller $CState$ and $CData$ which we will discuss in the next section.}
%\QIN{
%Another possible encoding of such ``Enable'' operator, called $EnableData$, uses a data-oriented style, with a single state controller $CData$, and a state-variable $v$, with values in $\{0,1\}$. As there is no more constructor required, we can use the same algebra presentation as $EnableState$ in this case.
%}
\end{example}

@@

\begin{example}[PLTS of the Controller in the ``Enable'' Operator]
\ERIC{
The PLTS of the controller $CState$ from Example \ref{example:EnableOperator} is  $\langle S,s_0, \to\rangle$, where:
\begin{itemize}
	\item[$\bullet$] $S=\{0, 1\}$, $s_0=0$;
	\item[$\bullet$] $\to=\{0\xrightarrow{l}0, 0\xrightarrow{\delta}1, 1\xrightarrow{r}1\}$.
\end{itemize}
}
\QIN{
Featuring guards and assignments in the label of transitions, the PLTS of the controller $CData$ is $\langle S,s_0, \to\rangle$ where:
\begin{itemize}
	\item[$\bullet$] $S=\{0\}$, $s_0=0$, $vars(s_0)=\{v\}$, $\texttt{init}\ \{v:=0\}$;
	\item[$\bullet$] $\to=\{0\xrightarrow{l, [v=0]}0, 0\xrightarrow{\delta,  [v=0], \{v\leftarrow1\}}0, 0\xrightarrow{r, [v=1]}0\}$.
\end{itemize}
}
\ERIC{The PLTSs contained in $EnableStateCompLeft$ are similar to the PLTS $CState$, but their action constants have been renamed, ensuring disjoint state variable sets.}
\end{example}

@@

\begin{example}[pNet of the ``Enable'' Operator]\label{example:pNetEnable}
\ERIC{
  Consider the pNet $EnableState$ shown in Fig. \ref{schema:enable-pnets}, upper part. It has two holes $P$ and $Q$, and one PLTS $CState$. Note that our graphical syntax presents pNet nodes and synchronisation vectors as finite arrays, where the index sets follow the positions in the arrays.
  In this example we use \emph{synchronised actions}, to express actions that cannot be further synchronised, similar to the silent action $\tau$ of CCS. We denote them as any action expression but with the text underlined, as e.g. $\underline{\delta(x_1)}$. Such synchronised actions do not play any special role for defining the semantics of pNets, but as one can expect, will be crucial for defining weak equivalences. The pNet can be given as follow:
\begin{itemize}
	\item[$\bullet$] $\pNet=\{CState\}$;
	\item[$\bullet$] $J=\{P, Q\}$;
	\item[$\bullet$] $\symb{SV}=$ as listed on the right side of each pNet node.
        %%   \{\langle a_1, \text{-}, l\rangle\rightarrow a_1[\forall y1.a_1\neq \delta(y1)], 
	%% \langle \delta(x_1), acc(x_1), \delta\rangle\rightarrow\underline{\delta(x_1)},\\
	%% \langle \text{-}, a_2, r\rangle\rightarrow a_2[\forall y2.a_2\neq acc(y2)]\}$.
\end{itemize}
Consider the first  vector of $EnableState$, written as $\langle a_1, \text{-}, l\rangle\rightarrow a_1[\forall y1. a_1\neq \delta(y1)]$. It denotes a possible behavior of $P>>Q$, for any possible action of the hole $P$ that respects the vector predicate, that is any action that does not match $\delta(y1)$, for any value of the variable $y1$.
Remark that the data-oriented encoding, in pNet $EnableData$ has the same synchronisation vectors, upto alpha conversion. Only the controller $CState$ has been replaced by $CData$.
}
\end{example}

@@

%\QIN{The pNet of $EnableStateCompLeft$ presents a bigger tree with three levels containing a pNet as its node.}
\begin{example}[pNet of the Composition of ``Enable'' Operators]\label{example:pNetComposeEnable}
\ERIC{
  Now consider the pNet $EnableStateCompLeft$ in Fig. \ref{schema:enable-composed}; the processes $P$, $Q$ and $R$ are holes of the pNet.
It has been obtained by replacing the second hole in pNet $EnableState$ by another pNet representing $Q>>R$. Technicaly, the action constants in the PLTS $CState1$ and $CState2$ have been renamed to avoid clashes, and the variables in the vectors also given different names.
%%   Let $pNet_1$ be the sub-pNet that presents $Q>>R$. The pNet can be given as follow:
%% \begin{itemize}
%% 	\item[$\bullet$] $\pNet=\{pNet_1, CState1\}$;
%% 	\item[$\bullet$] $J=\{P\}$;
%% 	\item[$\bullet$] $\symb{SV}=\{\langle a_1, \text{-}, l1\rangle\rightarrow a_1[\forall y1.a_1\neq \delta(y1)], 
%% 	\langle \delta(x_1), acc(x_1), \delta\rangle\rightarrow\underline{\delta(x_1)},\\
%% 	\langle \text{-}, a_2, r1\rangle\rightarrow a_2[\forall y2.a_2\neq acc(y2)]\}.$
%% \end{itemize}
%% The sub-pNet $pNet_1$ is a pNet of a single ``Enable'' operator as the pNet in Example \ref{example:pNetEnable}. Actions in $pNet_1$ should be renamed to distinguish themselves from those already used in the pNet of $EnableStateCompLeft$.
}
\end{example}

@@

\ERIC{Each call to the SMT engine will return either SAT or UNSAT results, or may fail due to time bounds. Also, the accuracy of a `SAT' result may depend on the characteristics of the specific theory used to axiomatise the data domains and operators. Such theories may be intrinsically undecidable, or may not properly handled by the SMT engine (e.g. for recursive datatypes).
However, while building the open automaton, that is a symbolic representation of the pNet behavior, it is not logically incorrect to keep potentially unsatisfiable transitions, so we will treat undecidable results, and time bounds failures, as SAT. So the open automaton, may have more transitions and states that the theoretical minimum.}

\ERIC{So this is not a problem in itself when constructing the automaton, but a similar limitation will arise later when using open automata as the input of model checking or equivalence checking algorithms. Note that decidability results for these 
will be dependent on the decidability of the SMT theories used, as is proved e.g. in \cite{henrio:Forte2016} for bisimulation.}

@@

\QIN{As the removal of the unsatisfiable transitions leads to some transitions losing their pioneer transitions, }
the set $L1$ may contain unreachable transitions. To remove those transitions, the last step of the  
algorithm (lines 12-14) builds the final open automaton via a traversal of $L1$,
starting at the initial state $InitState(p)$ as defined in
page~\pageref{def-states}. More precisely, this is a residual
algorithm building a graph from the reachable transitions in $L1$. 
%and states are the set of target states of the transitions. 
Each state in the graph is decorated with a set of its local
variables, and a list of their possible assignments,
gathered during the traversal using the ``Post'' parts of the
OTs.

@@

In practice, this could be
expensive \QIN{as it needs several times of traversal of the whole set of open transitions}, and it is not incorrect to leave some UNSAT OTs in the
result. \QIN{From a  practical point of view, the unreachable transitions is usually a small part of result open transitions, very little will be gained to compute the fixed point. So we would only perform one step of iteration.} Further algorithms 
(e.g. model-checking or equivalence checking) will be able to use the
full information from the predicates and the assignment sets, even if
some spurious OTs are left here.

@@

The open automata generated for the two encodings of Enable in
Figure~\ref{schema:enable-pnets} are displayed in
Figure~\ref{schema:enable-OAs}. 
\QIN{The initial transition of $EnableData$ is transformed to the initialisation of the state variables, here is $\{v\leftarrow1\}$, for its open automata.}
Note that although these automata look very similar to the PLTS
controllers from Figure~\ref{schema:enable-pnets}, their nature is very
different: they represent the behavior of the whole system, including
the holes $P$ and $Q$. Their transitions do not rely on a particular
implementation of the controllers, but only encode the 
relations between the actions of the holes.
We have shown in \cite{henrio:Forte2016} 
\YX{that the two open automata are equivalent with respect to a notion of FH-bisimulation.}
%how to prove that they are bisimilar.

@@

Intuitively, application of the Failure Monitor architecture ensures
that, whenever a failure is registered in the operand component, the
system will be reset, unless a resumption is registered within
$\mathrm{Max}$ time units.  Formally, this statement is decomposed
into several properties, among which 
%
1)~the safety property \emph{``A reset is never asked for unless a failure occurs''} specified using CTL as
%
\begin{equation}
  \label{eq:FM:safety}
  %% \AG
  %% \bigl(\PortFail \rightarrow
  %% \EX \EU[]{\lnot \PortFail}{\PortReset}\bigr)
  \AW[]{\lnot \PortAsk}{\PortFail}
\end{equation}
%
and 2)~the liveness property \emph{``in the absence of an external
  reset, a reset can always be asked for after a non-transient
  failure''}:
%
\begin{equation}
  \label{eq:FM:liveness}
  %% \mathtt{AG}(\PortAsk \rightarrow \mathtt{AF}\ \PortReset).
  \AG \bigl(
  \PortFail \rightarrow \EU[]{(\lnot \PortReset \land \lnot \PortResume)}{\PortAsk}
  \bigr)
\end{equation}
%
Additional properties are provided in~\cite{Avocs-RR}.

>>>>

\begin{example}[The ``Enable'' Operator]\label{example:EnableOperator}
\ERIC{In Fig. \ref{schema:enable-pnets} we give two possible pNet  encodings of the ``Enable'' operator, \YX{which are  state-oriented and data-oriented, respectively.} 
The first one is called $EnableState$, and uses a controller $CState$ with three simple control actions, $l, r$ and $\delta$,  to synchronise the behaviour of $P$ and $Q$. Here we use a simple action algebra, containing two constructors $\delta(x)$ and $acc(x)$, for any possible data type of the variable $x$, corresponding to the LOTOS statements \texttt{exit} and \texttt{accept}. To construct the algebra presentation, we extend the triple $\APres=\langle\mathit{Sorts},\mathit{Constrs},\mathit{Ops}\rangle$ shown in Table \ref{table:BIPalgebra} with four new action constant constructors $l, r, \delta$ and $acc$.}

@@

\ERIC{The $EnableData$ pNet uses the same algebra presentation.}\QIN{The main difference between $EnableState$ and $EnableData$ lies in the construction of their controllers $CState$ and $CData$ that we will discuss in the next section.}
%\QIN{
%Another possible encoding of such ``Enable'' operator, called $EnableData$, uses a data-oriented style, with a single state controller $CData$, and a state-variable $v$, with values in $\{0,1\}$. As there is no more constructor required, we can use the same algebra presentation as $EnableState$ in this case.
%}
\end{example}

@@

\begin{example}[PLTS of the Controller in the ``Enable'' Operator]
\ERIC{
The PLTS of the controller $CState$ from Example \ref{example:EnableOperator} is  $\langle S,s_0, \to\rangle$, where:
\begin{itemize}
	\item[$\bullet$] $S=\{0, 1\}$, $s_0=0$;
	\item[$\bullet$] $\to=\{0\xrightarrow{l}0, 0\xrightarrow{\delta}1, 1\xrightarrow{r}1\}$.
\end{itemize}
}
\QIN{
%Featuring guards and assignments in the label of transitions, 
\YX{The PLTS of the controller $CData$ is $\langle S',s'_0, \to\rangle$} where:
\begin{itemize}
	\item[$\bullet$] \YX{$S'=\{0\}$, $s'_0=0$, $vars(s'_0)=\{v\}$;} % $\texttt{init}\ \{v:=0\}$;
	\item[$\bullet$] $\to=\{0\xrightarrow{l, [v=0]}0, 0\xrightarrow{\delta,  [v=0], \{v\leftarrow1\}}0, 0\xrightarrow{r, [v=1]}0\}$.
\end{itemize}
}
\ERIC{The PLTSs contained in $EnableStateCompLeft$ are similar to the PLTS $CState$, but their action constants have been renamed, ensuring disjoint state variable sets.}
\end{example}

@@

\begin{example}[pNet of the ``Enable'' Operator]\label{example:pNetEnable}
\ERIC{
  Consider the pNet $EnableState$ shown in Fig. \ref{schema:enable-pnets}, upper part. It has two holes $P$ and $Q$, and one PLTS $CState$. Note that our graphical syntax presents pNet nodes and synchronisation vectors as finite arrays, where the index sets follow the positions in the arrays.
  In this example we use \emph{synchronised actions}, to express actions that cannot be further synchronised, similar to the silent action $\tau$ of CCS. We denote them as any action expression but with the text underlined,  e.g. $\underline{\delta(x_1)}$. Such synchronised actions do not play any special role for defining the semantics of pNets, but as one can expect, they will be crucial for defining weak equivalences. The pNet can be given as follow:
\begin{itemize}
	\item[$\bullet$] $\pNet=\{CState\}$;
	\item[$\bullet$] $J=\{P, Q\}$;
	\item[$\bullet$] $\symb{SV}=$ as listed on the right side of each pNet node.
        %%   \{\langle a_1, \text{-}, l\rangle\rightarrow a_1[\forall y1.a_1\neq \delta(y1)], 
	%% \langle \delta(x_1), acc(x_1), \delta\rangle\rightarrow\underline{\delta(x_1)},\\
	%% \langle \text{-}, a_2, r\rangle\rightarrow a_2[\forall y2.a_2\neq acc(y2)]\}$.
\end{itemize}
Consider the first  vector of $EnableState$, written as $\langle a_1, \text{-}, l\rangle\rightarrow a_1[\forall y1. a_1\neq \delta(y1)]$. It denotes a possible behavior of $P>>Q$, for any possible action of the hole $P$ that respects the vector predicate, that is, any action that does not match $\delta(y1)$, for any value of the variable $y1$.
Remark that the data-oriented encoding in pNet $EnableData$ has the same synchronisation vectors, up to the usual alpha conversion. Only the controller $CState$ has been replaced by $CData$.
}
\end{example}

@@

%\QIN{The pNet of $EnableStateCompLeft$ presents a bigger tree with three levels containing a pNet as its node.}
\begin{example}[pNet of the Composition of ``Enable'' Operators]\label{example:pNetComposeEnable}
\ERIC{
  Now consider the pNet $EnableStateCompLeft$ in Fig. \ref{schema:enable-composed}; the processes $P$, $Q$ and $R$ are three holes of the pNet.
It has been obtained by replacing the second hole in pNet $EnableState$ by another pNet representing $Q>>R$. Technically, the action constants in the PLTS $CState1$ and $CState2$ have been renamed to avoid clashes, and the variables in the vectors are also given different names.
%%   Let $pNet_1$ be the sub-pNet that presents $Q>>R$. The pNet can be given as follow:
%% \begin{itemize}
%% 	\item[$\bullet$] $\pNet=\{pNet_1, CState1\}$;
%% 	\item[$\bullet$] $J=\{P\}$;
%% 	\item[$\bullet$] $\symb{SV}=\{\langle a_1, \text{-}, l1\rangle\rightarrow a_1[\forall y1.a_1\neq \delta(y1)], 
%% 	\langle \delta(x_1), acc(x_1), \delta\rangle\rightarrow\underline{\delta(x_1)},\\
%% 	\langle \text{-}, a_2, r1\rangle\rightarrow a_2[\forall y2.a_2\neq acc(y2)]\}.$
%% \end{itemize}
%% The sub-pNet $pNet_1$ is a pNet of a single ``Enable'' operator as the pNet in Example \ref{example:pNetEnable}. Actions in $pNet_1$ should be renamed to distinguish themselves from those already used in the pNet of $EnableStateCompLeft$.
}
\end{example}

@@

\ERIC{Each call to the SMT engine will return either SAT or UNSAT results, or may fail due to time bounds. Also, the accuracy of a `SAT' result may depend on the characteristics of the specific theory used to axiomatise the data domains and operators. Such theories may be intrinsically undecidable, or may not properly handled by the SMT engine (e.g. for recursive datatypes).
However, while building the open automaton, that is a symbolic representation of the pNet behavior, it is not logically incorrect to keep potentially unsatisfiable transitions, so we will treat undecidable results, and time bounds failures, as SAT. So the open automaton may have more transitions and states than the theoretical minimum.}

\ERIC{This is not a problem in itself when constructing the automaton, but a similar limitation will arise later when using open automata as the input of model checking or equivalence checking algorithms. Note that decidability results for these 
will be dependent on the decidability of the SMT theories used, as is proved e.g. in \cite{henrio:Forte2016}, for bisimulations.}

@@

\QIN{As the removal of the unsatisfiable transitions \YX{may entail some transitions without proceeding} transitions, }
the set $L1$ might contain unreachable transitions. To remove those transitions, the last step of the  
algorithm (lines 12-14) builds the final open automaton via a traversal of $L1$,
starting at the initial state $InitState(p)$ as defined in
page~\pageref{def-states}. More precisely, this is a residual
algorithm building a graph from the reachable transitions in $L1$. 
%and states are the set of target states of the transitions. 
Each state in the graph is decorated with a set of its local
variables, and a list of their possible assignments,
gathered during the traversal using the ``Post'' parts of the
OTs.

@@

In practice, this could be
expensive \QIN{as it needs several times of traversal of the whole set of open transitions}, and it is not incorrect to leave some UNSAT OTs in the
result. \QIN{\YX{Empirically, the unreachable transitions constitute  a small part of the ultimate} open transitions, very little will be gained to compute the fixed point. So we would only perform one step of iteration.} Further algorithms 
(e.g. model-checking or equivalence checking) will be able to use the
full information from the predicates and the assignment sets, even if
some spurious OTs are left here.

@@

The open automata generated for the two encodings of Enable in
Figure~\ref{schema:enable-pnets} are displayed in
Figure~\ref{schema:enable-OAs}. 
\QIN{The initial transition of $EnableData$ is transformed to the initialisation of the state variables, which is $\{v\leftarrow1\}$, for its open automata.}
Note that although these automata look very similar to the PLTS
controllers from Figure~\ref{schema:enable-pnets}, their nature is very
different: they represent the behavior of the whole system, including
the holes $P$ and $Q$. Their transitions do not rely on a particular
implementation of the controllers, but only encode the 
relations between the actions of the holes.
We have shown in \cite{henrio:Forte2016} 
\YX{that the two open automata are equivalent with respect to a notion of FH-bisimulation.}
%how to prove that they are bisimilar.

@@

Intuitively, application of the Failure Monitor architecture ensures
that, whenever a failure is registered in the operand component, the
system will be reset, unless a resumption is registered within
$\mathrm{Max}$ time units.  Formally, this statement is decomposed
into several properties, among which 
%
1)~the safety property \emph{``A reset is never asked for unless a failure occurs''} specified using CTL as
%
\begin{equation}
  \label{eq:FM:safety}
  %% \AG
  %% \bigl(\PortFail \rightarrow
  %% \EX \EU[]{\lnot \PortFail}{\PortReset}\bigr)
  \AW[]{\lnot \PortAsk}{\PortFail}
\end{equation}
%
and 2)~the liveness property \emph{``in the absence of an external
  reset, a reset can always be asked for after a non-transient
  failure''}:
%
\begin{equation}
  \label{eq:FM:liveness}
  %% \mathtt{AG}(\PortAsk \rightarrow \mathtt{AF}\ \PortReset).
  \AG \bigl(
  \PortFail \rightarrow \EU[]{(\lnot \PortReset \land \lnot \PortResume)}{\PortAsk}
  \bigr)
\end{equation}
%
Additional properties are provided in~\cite{Avocs-RR}. \YX{Once we have generated the open automaton of the architecture, such properties can be verified by model checking tools, which is beyond the scope of the current work.}
